---
title: "Estimating fitness using BiSSeCO"
author: "Fabricia F. Nascimento"
date: "`r Sys.Date()`"
output:
  bookdown::html_vignette2:
  #bookdown::pdf_book:
    toc: TRUE
    fig_caption: TRUE
    citation_package: biblatex
bibliography: bib/tutorials.bib
pkgdown:
  as_is: true
fontsize: 12pt
vignette: >
  %\VignetteIndexEntry{Estimating fitness using BiSSeCO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(magrittr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(kableExtra.html.bsTable = TRUE)
```

# Introduction

This vignette will demonstrate how to to estimate within-host and between-host 
transmission fitness using the package `musseco` and the BiSSeCo
(Binary-State Speciation and extinction Coalescent) model.


# Basic requirements
This vignette assumes that you have the following packages already installed:

* [musseco](https://github.com/emvolz/musseco): implements the likelihood 
function to estimate fitness.
* [ape](https://github.com/emmanuelparadis/ape): used for manipulation of 
phylogenetic trees.
* [TiPS](https://cran.r-project.org/web/packages/TiPS/vignettes/TiPS.html): used
to simulate a phylogenetic tree according to an epidemiological model.
* [ggplot2](https://ggplot2.tidyverse.org/): used to visualize the results.


## Load the necessary packages:
```{r message = FALSE}
  library(musseco)
  library(TiPS)
  library(ape)
  library(ggplot2)
```


# Simulated phylogenetic tree
To illustrate how we can use the R package `musseco` to estimate relative fitness,
we will use a tree that was simulated with the R package `TiPS`. 

This tree can be load as:

```{r}
tr <- readRDS(system.file("extdata/tips_params_1_rep_2.rds", package = "musseco"))
```


## Epidemiological model
This tree was simulate in a way that we know the true values of within-host and 
between-host replicative fitness.

The epidemiological model we used are described by the following equations,
describing how the number of individuals carrying the ancestral ($Y_a$) and variant
($Y_v$) sequences changes with time.

$\dot{Y_a} = \beta Y_a (1 - \dfrac{N}{K}) + \mu_{va}Y_v - \mu_{av}Y_a - \gamma Y_a$

$\dot{Y_v} = \beta(1+s)Y_v (1 - \dfrac{N}{K}) - \mu_{va}Y_v + \mu_{av}Y_a - \gamma Y_v$


where the rate of new infections was defined as logistic growth controlled by the 
parameter $\beta$ for $Y_a(t)$ and $\beta(1+s)$, where $s$ is the selection coefficient, 
for $Y_v(t)$. 

The carry capacity was defined by the parameter $K$ and was fixed at 
10,000 individuals.

The rate $\mu_{av}$ defined how individuals carrying the ancestral state $Y_a(t)$ 
mutated to the $Y_v(t)$.
The rate $\mu_{va}$ defined how individuals carrying the variant mutation 
reversed to the ancestral state, and this is also the molecular clock rate of 
evolution.

The rate $\gamma$ denoted the mortality rate.

In this vignette will not focus on how this tree was simulated,
but the scripts to simulate the tree can be found [here](https://github.com/thednainus/musseco_simulations/blob/main/coalescent_sim/TiPS_simulations/simulate_TiPS_trees.R).

Table 1 shows the parameter values used to simulate the tree


```{r table1, echo = FALSE}
Parameter = c("Growth rate",
              "Carrying capacity",
              "Mortality rate",
              "Transition rate from $Y_v$ to $Y_a$",
              "Transition rate from $Y_a$ to $Y_v$",
              "Selection coefficient",
              "Between-host replicative fitness",
              "Within-host replicative fitness") 
Symbol = c("$\\beta$",
           "$K$",
           "$\\gamma$",
           "$\\mu_{va}$",
           "$\\mu_{av}$",
           "$s$",
           "$\\omega$",
           "$\\alpha$") 
Values = c("0.216",
           "10,000",
           "1/10.2",
           "0.0016",
           "0.00176",
           "-0.1",
           "0.9",
           "1.1") 
data_list = data.frame(Parameter, Symbol, Values)

kableExtra::kbl(data_list, booktabs = TRUE, escape = FALSE, 
    caption = "Parameter symbols and values") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))

```

Based on this parameter values, we know the true value of $\alpha$ and $\omega$.

Note that $\alpha = \mu_{av}/\mu_{va}$ and $\omega = 1 + s$.


# Estimation of parameter values using fitbisseco

To estimate the within-host and between-host replicative fitness using the 
simulated phylogenetic tree, we will use the function _fitbisseco_ from the 
`musseco` package.

Note that we need to define which tips of the phylogenetic tree carries the 
variant state.

```{r}
isvariant <- grepl( tr$tip.label, pattern = 'V_' )
isvariant <- setNames(isvariant, tr$tip.label)

```

We also need to define the value of mu, which is the molecular clock
rate of evolution. In our example, we used the rate for HIV-1 subtype C 
[@Patino-Galindo_2017] as we were modelling HIV-1 dynamics.

We also need to define the generation time (Tg).
The generation time can be defined as $1/\gamma$. 

Based on our Table 1, we know the values of mu and gamma.

```{r}
# gamma value used in our epidemiological model
#Tg = 1/gamma
gamma <- 1/10.2
# mu (molecular clock) value used in our epidemiological model
mu <- 0.0016
```



Now we can run the _fitbisseco_ function using the following command:

```{r eval = FALSE}
fb_au <- fitbisseco( tr,
                     isvariant,
                     Tg = 1/gamma,
                     mu = mu,
                     Net = NULL,
                     theta0 = log(c(2,.75,1/2)),
                     augment_likelihood = TRUE,
                     optim_parms = list(),
                     mlesky_parms = list(tau = NULL,
                                         tau_lower = 0.1,
                                         tau_upper = 1e7,
                                         ncpu = 1, 
                                         model = 1 ) )
```

Note the following options:

* `tr` is the `ape::phylo` object, and this must be a time-scaled phylogeny.
* `isvariant` vector of type boolean with lentgh equal to the number of tips in
   the phylogenetic tree. Each element is `TRUE` if the corresponding element in tr$tip.label 
   is a variant type.
* `Tg` is the generation time, i.e. the mean time elapsed between generations.
* `mu` is the molecular clock rate of evolution;
* `Net` (default `NULL`): but user can also provide a matrix with two columns 
   giving the effective population size. If NULL, the effective population 
   size will be computed with the [mlesky](https://github.com/emvolz-phylodynamics/mlesky) 
   R package. The first column should be time since some point in the past, and 
   the second column should be an estimate of the effective population size. 
   Time zero should correspond to the root of the tree.
* `theta0`: Initial guess of (log) parameter values: alpha, omega, and Ne scale. 

For details on the other parameters see `?musseco::fitbisseco`.

But user can also provide additional parameter values to use with optim when 
fitting the model by maximum likelihood, mlskygrid function from the [mlesky](https://github.com/emvolz-phylodynamics/mlesky)
R package and the [colik](https://emvolz-phylodynamics.github.io/phydynR/reference/colik.html) 
function from the [phydynR](https://emvolz-phylodynamics.github.io/phydynR/) package.

In addition, the _fitbisseco_ function implements two types of likelihoods: 
the ‘coalescent’ and the ‘augmented’ likelihood.
Both methods are based on the conditional density of a genealogy given epidemic 
and demographic parameters. 
The coalescent likelihood (by setting augment_likelihood = FALSE) relies solely 
on the phylogenetic tree to infer the fitness parameters. 
The augmented likelihood combines in a simple sum, the coalescent likelihood with
a binomial likelihood of sampling variants or ancestral types under the assumption 
of random sampling and mutation-selection balance. 
The augmented likelihood is the default option in the _fitbisseco_ function.

## Obtaining the estimated parameter values

The fitbisseco function will take a few minutes to run.
You can load the fit here

```{r}
load(system.file("extdata/fit.rda", package = "musseco"))

fb_au
```

We can now plot alpha and omega in respect to the true values listed in Table 1.
The true values was those used to simulate the phylogenetic tree.

First we need to get the confidence interval as data.frame.

```{r}
# threshold for approximate CIs 
MAXCOEFVAR <- 1.5

#variable names  
vnames <-  c('alpha', 'omega', 'yscale')
#the best values for the parameters found using optim
best_values <- coef(fb_au)[vnames]

# create a dataframe with the confidence interval
odf = as.data.frame(best_values)
odf$lower_bound <- exp( log(coef(fb_au)[vnames])-fb_au$err*1.96 )
odf$upper_bound <- exp( log(coef(fb_au)[vnames])+fb_au$err*1.96 )
odf$upper_bound[ fb_au$err > MAXCOEFVAR ] <- Inf
odf <- round( odf, 3 )

#add the true values to data.frame
odf["true_values"] <- c(1.1, 0.9, NA)
odf["Variable"] <- c("alpha", "omega", NA)
odf$Variable <- as.factor(odf$Variable)
```

Plot we can plot the values for alpha and omega using ggplot.

```{r figure1, fig.asp = 0.8, fig.width = 6, fig.cap = "Bar plots showing the confidence interval for estimates of alpha and omega. Blue dot is the true value", echo = TRUE}
  ggplot(odf[1:2,], aes(x = Variable)) +
  geom_point(aes(y = best_values), colour = "black", size = 3) +
  geom_point(aes(y = true_values), colour = "blue", size = 3) +
  geom_errorbar(aes(ymax = upper_bound, ymin = lower_bound, 
                    width = 0.5), colour = "black") +
  facet_wrap( ~ Variable, scales = "free") +
  theme_bw() +
  xlab("True value") +
  ylab("Confidence Interval") +
  theme(text = element_text(size = 20), legend.position = "bottom")

```

# References